Automated Lensing Learner - I: An automated Strong Lensing Identification Pipeline

### 2017-03-23 ms.tex
Added grid search sample results for discussion

% Classical machine learning methods
Pattern recognition methods have transitioned to using ``machine
learning'' algorithms in place of manual cutoffs.  With machine
learning, we can train a model to separate a dataset into a known set
of classes, such as ``lensing systems'' and ``non-lensing
systems''. However, many classic machine learning algorithms do not
work well in image space, i.e. directly using the raw pixel values of
the image, but first require a process of \em{feature engineering}.
{\em Feature engineering} uses domain expertise 
to extract variables
that are more directly related to the classification task at hand.  An
example would be the application of an edge detector.  The optimal
weights and cutoffs for these derived variables that are used to
determine the class label of an image are found automatically by the
algorithm.

@@

\begin{table}
\caption{Grid Search of Pipeline Parameters}
\begin{center}
\begin{tabular}{cccccc}
\hline \\ [-0.2cm]
N$_{orient}$ & Pixels/Cell & Cells/Block & C$_{reg}$ & HOG size & Score \\ [0.2cm]
\hline \\ [-0.2cm]
\multicolumn{6}{c}{(a) HST-like data} \\ [0.2cm]
\hline \\ [-0.2cm]
9 & (4, 4) & (3, 3) & 50. & &$0.76181\pm0.00491$\\ [0.2cm]
9 & (4, 4) & (1, 1) & 50. & &$0.83403\pm0.00547$\\ [0.2cm]
9 & (8, 8) & (3, 3) & 50. & &$0.85417\pm0.00851$\\ [0.2cm] 
9 & (8, 8) & (1, 1) & 50. & &$0.86389\pm0.00687$\\ [0.2cm] 
9 & (16, 16) & (1, 1) & 50. & & $0.88542\pm0.00680$ \\ [0.2cm]
9 & (16, 16) & (3, 3) & 10. & &$0.89306\pm0.00687$ \\ [0.2cm]
8 & (16, 16) & (3, 3) & 50. & &$0.90000\pm0.00170$ \\ [0.2cm]
9 & (16, 16) & (3, 3) & 50. & &$0.90139\pm0.00804$ \\ [0.2cm]
\hline \\ [-0.2cm]
\multicolumn{6}{c}{(b) LSST-like data} \\ [0.2cm]
\hline \\ [-0.2cm]
& & & & &\\
& & & & &\\ [0.2cm]
\hline
\end{tabular}
\end{center}
\label{tab:gridsearch}
\tablecomments{Panel (a) shows the results of a grid search across a
  range of HOG parameters and regression coefficient \wording{with
    training and test sets of size 1440 with the HST-like mock
    dataset.}  Panel (b) shows the corresponding results for training
  and test sets of \wording{size 5760 with the LSST mock dataset.}  We
  explore different HOG parameters in each dataset due to resolution
  and image size differences.}
\end{table}

@@

Finally, the resolution of the overall image will also limit the
additional information that an increase in $N_\text{orient}$ will
provide.  \todo{From the grid search, the best case number of orientations
for each dataset is
$N_\text{orient,HST}=9$, $N_\text{orient,LSST10}=3$, and
  $N_\text{orient,LSST1}=$.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
\begin{center}
\includegraphics[width=2\columnwidth]{figures/tiles/compilations/hog_telescope_compilation.pdf}
\caption{Left to right show example image transforms of mock images
  from Figure~\ref{fig:mockimages} with a visualized histogram of
  oriented gradients.  The image transform picks up edge features,
  with arc features showing up as edges across radial orientations.
  \wording{Each of the oriented gradients within a cell is
    color-coded by magnitude, and represented as a line in the
    direction perpendicular to that gradient.}  The actual extracted
  features fill a one-dimensional feature vector comprised of the
  magnitudes of each of the oriented edges within the visualized
  cells.}\label{fig:hogimages}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%% \caption{Top row: Visualized HOG feature vectors of the sample
%%   lensed HST-like image from Figure~\ref{fig:mockimages}.  Left:
%%   The corresponding HOG visualization with N$_\text{orient}=5$,
%%   (32, 32) pixels per cell, and (1, 1) cells per block.  Right: The
%%   HOG visualization of the same mock image with
%%   N$_\text{orient}=9$, (16, 16) pixels per cell, and (1, 1) cells
%%   per block.  The brightest lines align with prominent edges.
%%   Note, while fewer pixels per cell will better resolve the arc
%%   feature, the ``edge'' corresponding to the arc is less prominent.
%%   Bottom row: Visualized HOG feature vectors for the LSST-like
%%   single epoch and stacked images with the best scoring HOG
%%   parameterization of N$_\text{orient}=4$, (3, 3) pixels per cell,
%%   and (3, 3) cells per block.  The single epoch image (left) does
%%   not have significant enough edge features to emerge in the HOG,
%%   but the stacked image (right) does.}  \end{center} \end{figure*}
%%   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Figure~\ref{fig:trainsizeLSST10} shows how the AUC and the optimal
logistic regression coefficient depend on the log of the size of the
training set for both the LSST10 data in solid blue lines, and the
nLSST10 data in dashed blue lines.  The AUC for models trained on the
LSST10 data improves almost linearly with the log of the training set
size, increasing from AUC=0.705 to AUC=0.788 when the train size is
increased from 500 lens/non-lens images to 8000 lens/non-lens images.
However, for the nLSST10 data, where the lens galaxy has been removed
from the images, the improvement is less dramatic.  With the same
increase in training size, the AUC for nLSST10 changes from just below
0.77 to just below 0.78.

In solid red, we see that the optimal value for $C_\text{LogReg}$ for
LSST10 roughly scales logarithmically with the log of the train size,
with an exception of the data point corresponding to train size of
2000.  Generally, a larger training set allows for an increase in
model complexity without reducing its ability to generalize.  This is
also true for the optimal $C_\text{LogReg}$ dependence on the number
of training images in the nLSST10 data.  But, the required complexity
is systematically less than for the LSST10 images.

@@

@@

% Train time
Figure~\ref{fig:traintimeLSST10} shows the train time for LSST10 and
nLSST10 as a function of the size of the training set for the optimal
regularization parameter for that subset of the training data.  Each model
uses features extracted with the same HOG parameterization from the
grid search and the optimal regularization parameter for that subset of
the training data.  The train time of a given model generally increases for
increasing regularization parameter.  For the subset of train size
$N_\text{train}=2000$ in LSST10, the optimal regularization parameter
happened to be $C_{LogReg}=100$, whereas it was $C_{LogReg}=500$ for
the subset of trainsize 1000, and $C_{LogReg}=1000$ for the subset of
trainsize 4000.  This makes the train time increase at train size
$N_\text{train}=2000$ for LSST10 less dramatic than the average
log-log slope of approximately 2.

@@

% AUC vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/trainsize/LSST10_with_nLSST10.pdf}
\caption{Solid (dotted) blue line: AUC for models with varying size of
  the training set for LSST10 (nLSST10).  Solid (dotted) red line:
  Optimal regularization parameter for training on the train size for
  LSST10 (nLSST10). The improvement of AUC scales linearly with the
  log of the training set size, but train time roughly scales
  logarithmically.  Note, LSST10 requires more model complexity to
  exceed the performance of nLSST10.}\label{fig:trainsizeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/traintime/LSST10.pdf}
\caption{Solid (dotted) blue line: Train time for models with varying
  size of the training set for LSST10 (nLSST10).  Train time roughly
  scales logarithmically.  Note, LSST10 requires more model complexity
  to exceed the performance of nLSST10, and therefore requires more
  training time for continual increase in
  performance.}\label{fig:traintimeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Augmenting dataset
To augment our training set, we rotated each image in the set by
multiples of $90^o$.  Since our feature extraction method of HOG is
not rotationally invariant, augmenting our data by a factor of four
naturally optimizes the use of available training data.  This has an
equivalent improvement to the study illustrated in
Figure~\ref{fig:trainsizeLSST10}.

@@

% Summary figure
Figure~\ref{fig:rotation_test} also summarizes the best-case results
of our models trained on our entire 10,000 training sets, and tested
on our holdout 1,000 test set.  Recall, however, that we expect model
performance on images containing the lens galaxy to improve further
with larger training sets (see Figure~\ref{fig:trainsizeLSST10}).

@@

Figure~\ref{fig:ROCsamplesLSST10} illustrates four images from each
paradigm for the hold-out test set of LSST10.  The trained model used
the entire 10,000 image training set.  The left two columns show lens
systems, and the right two columns show non-lens systems.  The
annotation in the top left corner of each images shows the score,
\todo{and the letters in the bottom right label the paradigm of
  performance.}

@@

% Some dependence on redshift and velocity dispersion, which is folded
% into magnitude
In our mock HST sample, all but one of our lensing systems with
magnifications $\lsim7$ will be successfully identified for
classification threshold scores $\gsim0.5$.  These highly magnified
systems typically have lenses that are at higher redshifts of
$z\gsim0.5$ (lighter colors in bottom row) and/or lens galaxies with
velocity dispersions higher than $\sigma_v\lsim230.$~km/s (darker
colors in right figure).  Note that lensing galaxies with smaller
velocity dispersions are less massive and therefore have a smaller
efficiency of lensing cross section.  The smaller efficiency means
that a background galaxy is less likely to be strongly lensed with
high magnification.  Also, due to hierarchical structure formation,
galaxies are less massive at higher redshifts, so the trends of model
performance with these three parameters are somewhat degenerate with
one another.

>>>>

% Classical machine learning methods
Pattern recognition methods have transitioned to using ``machine
learning'' algorithms in place of manual cutoffs.  With machine
learning, we can train a model to separate a dataset into a known set
of classes, such as ``lensing systems'' and ``non-lensing
systems''. However, many classic machine learning algorithms do not
work well in image space, i.e. directly using the raw pixel values of
the image, but first require a process of {\em feature engineering}.
{\em Feature engineering} uses domain expertise to extract variables
that are more directly related to the classification task at hand.  An
example would be the application of an edge detector.  The optimal
weights and cutoffs for these derived variables that are used to
determine the class label of an image are found automatically by the
algorithm.

@@

\begin{table}
\caption{Grid Search of Pipeline Parameters}
\begin{center}
\begin{tabular}{cccccc}
\hline \\ [-0.2cm]
N$_{orient}$ & Pixels/Cell & Cells/Block & C$_{reg}$ & Score \\ [0.2cm]
\hline \\ [-0.2cm]
\multicolumn{6}{c}{(a) HST-like data} \\ [0.2cm]
\hline \\ [-0.2cm]
9 & (16, 16) & (4, 4) & 10. & $0.90141\pm 0.00658$ \\ [0.2cm]
9 & (24, 24) & (3, 3) & 10. & $0.89266\pm 0.01033$ \\ [0.2cm]
9 & (16, 16) & (1, 1) & 10. & $0.87484\pm 0.00671$ \\ [0.2cm]
7 & (16, 16) & (4, 4) & 10. & $0.90031\pm 0.00956$ \\ [0.2cm]
5 & (16, 16) & (4, 4) & 10. & $0.89453\pm 0.00840$ \\ [0.2cm]
9 & (32, 32) & (4, 4) & 10. & $0.85828\pm 0.01004$ \\ [0.2cm]
9 & (32, 32) & (4, 4) & 50. & $0.87344\pm 0.01267$ \\ [0.2cm]
\hline \\ [-0.2cm]
\multicolumn{6}{c}{(b) LSST-like data} \\ [0.2cm]
\hline \\ [-0.2cm]
3 & (5, 5) & (4, 4) & 100. & $0.64328\pm 0.00769$ \\ [0.2cm]
4 & (5, 5) & (4, 4) & 100. & $0.62469\pm 0.00485$ \\ [0.2cm]
3 & (5, 5) & (2, 2) & 100. & $0.63281\pm 0.00869$ \\ [0.2cm]
5 & (5, 5) & (4, 4) & 100. & $0.62734\pm 0.00800$ \\ [0.2cm]
3 & (3, 3) & (4, 4) & 100. & $0.59109\pm 0.00313$ \\ [0.2cm]
3 & (3, 3) & (2, 2) & 100. & $0.57297\pm 0.00508$ \\ [0.2cm]
4 & (4, 4) & (3, 3) & 100. & $0.63125\pm 0.01227$ \\ [0.2cm]
\hline
\end{tabular}
\end{center}
\label{tab:gridsearch}
\tablecomments{Panel (a) shows a subsample of the results of a grid
  search for HST across a range of HOG parameters and regression
  coefficient. Panel (b) shows a subsample of the results of a grid
  search for LSST10. Each use a data set of size $2\times8000$ for
  cross validation to get the average scores and standard deviation.
  We explore different HOG parameters in each dataset due to
  resolution and image size differences.}
\end{table}

@@

Finally, the resolution of the overall image will also limit the
additional information that an increase in $N_\text{orient}$ will
provide.  From the grid search, the best case number of orientations
for each dataset is $N_\text{orient,HST}=9$,
$N_\text{orient,LSST10}=3$, and $N_\text{orient,LSST1}=5$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure*}[t]
\begin{center}
\includegraphics[width=2\columnwidth]{figures/tiles/compilations/hog_telescope_compilation.pdf}
\caption{Left to right show example image transforms of mock images
  from Figure~\ref{fig:mockimages} with a visualized histogram of
  oriented gradients.  The image transform picks up edge features,
  with arc features showing up as edges across radial orientations.
  Each of the oriented gradients within a cell is color-coded by
  magnitude, and represented as a line in the direction perpendicular
  to that gradient.  The actual extracted features fill a
  one-dimensional feature vector comprised of the magnitudes of each
  of the oriented edges within the visualized
  cells.}\label{fig:hogimages}
\end{center}
\end{figure*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%% \caption{Top row: Visualized HOG feature vectors of the sample
%%   lensed HST-like image from Figure~\ref{fig:mockimages}.  Left:
%%   The corresponding HOG visualization with N$_\text{orient}=5$,
%%   (32, 32) pixels per cell, and (1, 1) cells per block.  Right: The
%%   HOG visualization of the same mock image with
%%   N$_\text{orient}=9$, (16, 16) pixels per cell, and (1, 1) cells
%%   per block.  The brightest lines align with prominent edges.
%%   Note, while fewer pixels per cell will better resolve the arc
%%   feature, the ``edge'' corresponding to the arc is less prominent.
%%   Bottom row: Visualized HOG feature vectors for the LSST-like
%%   single epoch and stacked images with the best scoring HOG
%%   parameterization of N$_\text{orient}=4$, (3, 3) pixels per cell,
%%   and (3, 3) cells per block.  The single epoch image (left) does
%%   not have significant enough edge features to emerge in the HOG,
%%   but the stacked image (right) does.}  \end{center} \end{figure*}
%%   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% LSST10 and nLSST10 behavior
Figure~\ref{fig:trainsizeLSST} shows how the AUC depends on the log of
the size of the training set for both the LSST10 (LSST1) data in the
solid blue (red) line, and the nLSST10 (nLSST1) data in dashed blue
(red) line.  The AUC for models trained on the LSST10 data improves
almost linearly with the log of the training set size, increasing from
AUC=0.705 to AUC=0.788 when the train size is increased from
$2\times500$ lens/non-lens images to $2\times8000$ lens/non-lens
images.  However, for the nLSST10 data, where the lens galaxy has been
removed from the images, the improvement is less dramatic.  With the
same increase in training size, the AUC for nLSST10 changes from just
below 0.77 to just below 0.78.

@@

% LSST1 and nLSST1 behavior
In red, we see an analogous improvement for LSST1 and nLSST1.  Here,
the crossover happens at train size of $2x1000$, where an increase in
larger training data set size does not help improve the AUC for nLSST1.

@@

% Train time dependence
We illustrate the dependence of train time on both the size of the
training set and model complexity in Figures~\ref{fig:traintimeLSST10}
and \ref{fig:traintimeLSST1}.  In red,
Figure~\ref{fig:traintimeLSST10} shows that the optimal values for
$C_\text{LogReg}$.  For LSST10, $C_\text{LogReg}$ roughly scales
logarithmically with the log of the train size, with an exception of
the data point corresponding to train size of 2000.  Generally, a
larger training set allows for an increase in model complexity without
reducing its ability to generalize.  This is also true for the optimal
$C_\text{LogReg}$ dependence on the number of training images in the
nLSST10 data.  But, the required complexity is systematically less
than for the LSST10 images.

% Train time
In blue, Figure~\ref{fig:traintimeLSST10} also shows the train time
for LSST10 and nLSST10 as a function of the size of the training set
for the optimal regularization parameter for that subset of the
training data.  Each model uses features extracted with the same HOG
parameterization from the grid search and the optimal regularization
parameter for that subset of the training data.  The train time of a
given model generally increases for increasing regularization
parameter.  For the subset of train size $N_\text{train}=2000$ in
LSST10, the optimal regularization parameter happened to be
$C_{LogReg}=100$, whereas it was $C_{LogReg}=500$ for the subset of
trainsize 1000, and $C_{LogReg}=1000$ for the subset of trainsize
4000.  This makes the train time increase at train size
$N_\text{train}=2000$ for LSST10 less dramatic than the average
log-log slope of approximately 2.

@@

% Train time behavior for LSST1 and nLSST1
\wording{For LSST1, Figure~\ref{fig:traintimeLSST1} shows the same
  relationships between train time and train size in blue, and the
  optimal regulariztion parameter (or model complexity) in red.
  Contrary to what we found when comparing LSST10 to nLSST10, nLSST1
  requires more model complexity than LSST1.  Recall that LSST1 and
  nLSST1 correspond to single epoch simulated exposures with the best
  seeing; these images exhibit better resolution but worse signal to
  noise.  The switch in required model complexity corresponds to a
  trade-off between features from the lens galaxy providing additional
  information or swamping the signal from a strong lensing arc.}

% AUC vs. regularization on larger train/test set with comparable
% grid search in parameterization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
\includegraphics[width=\columnwidth]{figures/trainsize/combined_LSST10_with_nLSST10.pdf}
\caption{Solid (dotted) blue line: AUC for models with varying size of
  the training set for LSST10 (nLSST10).  Solid (dotted) red line:
  Same for LSST1. The improvement of AUC scales roughly linearly with
  the log of the training set size.  However, LSST10 and LSST1 have a
  steeper improvement with data training size.  The performance of
  nLSST10 and nLSST1 models trained on smaller training data sets is
  better than the respective LSST10 and LSST1 models trained on the
  same size data, but the LSST10 and LSST1 models outperform with
  larger size training data set.}\label{fig:trainsizeLSST}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
  \includegraphics[width=\columnwidth]{figures/traintime/LSST10.pdf}}
\caption{Solid (dotted) blue line: Train time for models with varying
  size of the training set for LSST10 (nLSST10).  Train time roughly
  scales logarithmically with the train size, but the train time is
  also affected by model complexity.  Solid (dotted) red line: Best
  regularization parameter as a function of train size.  Note, LSST10
  requires more model complexity to exceed the performance of nLSST10
  (see blue solid and dotted lines in
  Figure~\ref{fig:trainsizeLSST}), and therefore requires more
  training time for continual increase in
  performance. }\label{fig:traintimeLSST10}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%            
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%                                        
\begin{figure}[t]
\begin{center}
  \includegraphics[width=\columnwidth]{figures/traintime/LSST1.pdf}}
\caption{Same as Figure~\ref{fig:traintimeLSST10}, but for LSST1.
  Here, nLSST1 requires more model complexity than LSST1.  Note, that
  LSST1 and nLSST1 have better seeing but reduced overall signal to
  noise than LSST10 and nLSST10.  This corresponds to better
  resolution, and therefore sharper edge features that prominently
  corespond to arcs.}\label{fig:traintimeLSST1}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Augmenting dataset
To augment our training set, we rotated each image in the set by
multiples of $90^o$.  Since our feature extraction method of HOG is
not rotationally invariant, augmenting our data by a factor of four
naturally optimizes the use of available training data.  This has an
equivalent improvement to the study illustrated in
Figure~\ref{fig:trainsizeLSST}.

@@

% Summary figure
Figure~\ref{fig:rotation_test} also summarizes the best-case results
of our models trained on our entire 10,000 training sets, and tested
on our holdout 1,000 test set.  Recall, however, that we expect model
performance on images containing the lens galaxy to improve further
with larger training sets (see Figure~\ref{fig:trainsizeLSST}).

@@

Figure~\ref{fig:ROCsamplesLSST10} illustrates four images from each
paradigm for the hold-out test set of LSST10.  The trained model used
the entire 10,000 image training set.  The left two columns show lens
systems, and the right two columns show non-lens systems.  The
annotation in the top left corner of each images shows the score.

@@

% Some dependence on redshift and velocity dispersion, which is folded
% into magnitude
In our mock HST sample, all but one of our lensing systems with
magnifications $\lsim7$ will be successfully identified for
classification threshold scores $\gsim0.5$.  These highly magnified
systems typically have lenses that are at higher redshifts of
$z\gsim0.5$ (lighter colors in bottom row) and/or lens galaxies with
velocity dispersions larger than $\sigma_v\gsim230.$~km/s (darker
colors in right figure).  Note that lensing galaxies with smaller
velocity dispersions are less massive and therefore have a smaller
efficiency of lensing cross section.  The smaller efficiency means
that a background galaxy is less likely to be strongly lensed with
high magnification.  Also, due to hierarchical structure formation,
galaxies are less massive at higher redshifts, so the trends of model
performance with these three parameters are somewhat degenerate with
one another.
